The importance of a complete statistical description of dependence between variables for ecology and related fields

### 2019-03-22 Paper.Rmd
finished moving most nonparam results to sup mat, the paper is shorter now

knitr::kable(tab_multivar_summary[1:9,], 
             format="latex", escape = F,
             caption.short = "Summary of results for multivariate datasets",
             caption = "Summary of results for multivariate datasets. Rows 3-9 of the table were computed for the non-independent pairs (row 2) only; row 10 was for all location pairs. A well-fitting location pair (row 6) was one for which the best-fitting copula had $p$-values $>0.01$ for both the Cramer-von Mises and Kolmogorov-Smirnov goodness of fit tests. \\label{tab_multivar_summary}",
             booktabs=T, linesep = "\\addlinespace",
             col.names = c(" ","Green spruce aphid abundance",
                        "Leaf-curling plum aphid first flight",
                        "Ceratium furca abundance","Methane flux")
             )%>%column_spec(2:5,width="2.5cm")%>%
             column_spec(1, width = "7cm")%>%kable_styling(latex_options = "basic",position = "center")
```

@@

Having demonstrated that non-N copula structure and asymmetric tail dependence are common
features of ecological data, we addressed Q2 by exploring via mathematical 
models two possible mechanisms of this structure which 
seem likely to operate in diverse ecological scenarios. The first mechanism 
is an extension of the well known Moran
effect, and was summarized conceptually in the Introduction. 
It explores the hypothesis that non-N copula structure and asymmetric tail dependence 
between environmental variables measured through time 
in two places may bring about similarly 
structured dependence between ecological variables that are influenced by the 
environmental variables.

Let $E_i(t)$ be an ecological variable 
measured in location $i$ ($i=1,2$).
We initially assumed the variable follows AR(1) dynamics within a location, 
$E_i(t+1)=\beta E_{i}(t)+ \sqrt{1-\beta^2} \epsilon_i(t)$. 
We used $\beta=0.5$. Here the $\epsilon_i(t)$ represent environmental noise; they are 
standard-normal random variables that are independent for distinct times, $t$, but 
that were made to exhibit different kinds of 
dependence across locations in different simulations (see below).
The ecological variable, $E$, is very general, and could represent, for instance, deviations
of a population density from a carrying capacity or average density, deviations of
total plant community biomass or other community variable from an average value,
or flux of a biogeochemical variable such as methane or CO$_2$..

@@

For each value of $b$, we simulated the model once using $g_1$ and once using $g_2$, for 5000 time
steps in both cases, retaining $E_1(t)$ and $E_2(t)$ for the final 500 time 
steps of each simulation. We applied our model selection algorithms and 
non-parametric statistics, which were previously applied to 
data to answer Q1, to the 
simulation outputs to discover if the model could
give rise to non-N copula structure and asymmetric tail dependence in the 
ecological variable.

@@

For strong noise and using our nonlinear model (methods for Q2), 
copula statistics, generally speaking, were approximately similar 
between noise and model
outputs; however, similarity was reduced relative to the AR(1) 
and weak-noise cases, and, for a few simulations, tail dependence 
asymmetry statistics
even had opposite signs for noise and model outputs 
(Figs \ref{SM-fig_Cause4copula_C_kend_sRicker_nsd_1}-\ref{SM-fig_Cause4copula_SJ_kend_sRicker_nsd_1}). For instance, using a C copula with a large Kendall $\tau$,
$\cor_l - \cor_u$ was slightly positive for noise, but slightly negative for 
population model outputs; and likewise for SG and SJ copulas. Dynamics were
strongly nonlinear for the model parameters we used, and nevertheless similarities
between noise and model-output copula structure, according to the statistics 
we used, were more noticable than discrepancies.
We again repeated the analysis with $r=1.3$. Discrepancies between noise and population
copula structure in this case were often glaring. Apparently noise of standard deviation 1
interacted very strongly with model nonlinearities when the model was in its overcompensatory 
regime.

Our model with environmental effects which differed in the
two tails could produce simulated ecological variables with visually apparent asymmetry 
of tail dependence and non-N copula structure, depending on the value of $b$ used. 
For $b=0.1$ (Fig. \ref{fig_ms_asym_b_0.1}) and $b=-0.1$ (Fig. \ref{SM-fig_ms_asym_b_-0.1}), 
for both $g_1$ and $g_2$,
asymmetric tail dependence and non-N copula structure were visually apparent.
Lower-tail (respectively, upper-tail) effects of the noise ($g_1$, resp., $g_2$) 
produced lower-tail (resp., upper-tail) dependence in the ecological variable.
Model selection results (Table \ref{SM-tab_summary_stat_asym_sens}) and
our nonparametric statistics (Tables \ref{SM-tab_stat_npa_asym}) 
statistically confirmed the visual result for $b=0.1, -0.1$. Results were more
complex for $b=0.9$ or $-0.9$, indicating that strongly autocorrelated (or strongly 
negatively autocorrelated dynamics) interact with the effect we have described.
Ecological-variable copulas were strongly non-N for $g_1$, but not for $g_2$ 
and asymmetries of tail dependence were not particularly noticable 
(Figs \ref{SM-fig_ms_asym_b_0.9}-\ref{SM-fig_ms_asym_b_-0.9} and Tables 
\ref{SM-tab_summary_stat_asym_sens}-\ref{SM-tab_stat_npa_asym}).

@@

We addressed Q3 using both models and data. An initial 
idea was presented in the Introduction that
the distribution (through time) of a spatially averaged quantity should be influenced by the 
nature of the dependence between the local quantities being averaged, including their
copula structure and tail dependence. For instance, suppose $E_i(t)$ are ecological
variables measured at locations $i=1,\ldots,N$ and times $t=1,\ldots,T$, and suppose 
the distribution and temporal dynamics of the spatial average $\sum_i E_i(t)/N$ is of
interest. The $E_i(t)$ could be, for instance, local abundances of a pest or 
exploited species, or local fluxes of a greenhouse gas. If  $E_i$ and $E_j$ are 
strongly right-tail dependent for all or most location pairs $i$ and $j$, then
exceptionally large values tend to occur at the same time in all or most locations.
This should tend to increase the skewness of the distribution through time of the
spatial average. Similarly, left-tail dependence between local variables
will tend to decreases skewness of the spatial mean. Positive skewness of the 
spatial mean corresponds to the "spikiness" of the spatial-mean time series. 
The spatial-mean time series and its skewness may be quantities of 
principal importance for 
cases such as pest or resource abundance or CO$_2$ or methane flux.

@@

We also examined whether the nature of the
tail dependence of an environmental variable across space could influence the 
extinction risk of a metapopulation. We hypothesized that environmental noises
exhibiting greater left-tail dependence across different habitat patches would 
cause higher metapopulation extinction risks because then bad years for the 
component populations would occur simultaneously in many patches,
reducing rescue effects. 
We provided a test of the conceptual validity of this hypothesis using a metapopulation
extension of the Lewontin-Cohen model,
$\vec{P}(t+1) = D \lambda(t) \vec{P}(t)$,
where $\vec{P}(t)$ is a vector of population densities, 
the $i^{th}$ component $P_i(t)$ representing
the density in the $i^{th}$ habitat patch ($i=1,\ldots,N$) at time $t$. 
The $N \times N$ matrix $\lambda(t)$ was diagonal with $i^{th}$ diagonal entry 
$\exp(r+\epsilon_i(t))$. Here $r$ is a growth rate; we used $r=0$. The
$\epsilon_i(t)$ represent environmental noises in each patch. They were standard-normally 
distributed in each patch, were independent through time, and showed the same 
spatial correlations (to within sampling error) for every simulation, 
but were made to exhibit right- or left-tail dependence between patches in
different simulations (Appendix \ref{SM-ext_risk_model}). The $N \times N$
matrix $D$ was a dispersal matrix. For simplicity, we considered habitat patches 
to be arranged in a line, with dispersal occurring either to nearest neighbors only,
or globally, in different simulations. The dispersal 
rate was $d$ (Appendix \ref{SM-ext_risk_model}). After each time step, 
if the population density in a patch was below $1$ (an extinction threshold), it was
set to $0$. We simulated the model 10000 times for each combination of parameters,
starting from initial populations $p_0 = 50$ in each patch, 
and calculated extinction risk after $25$ time steps.

>>>>

knitr::kable(tab_multivar_summary[1:9,], 
             format="latex", escape = F,
             caption.short = "Summary of results for multivariate datasets",
             caption = "Summary of results for multivariate datasets. Rows 3-9 of the table were computed for the non-independent pairs (row 2) only. A well-fitting location pair (row 6) was one for which the best-fitting copula had $p$-values $>0.01$ for both the Cramer-von Mises and Kolmogorov-Smirnov goodness of fit tests. \\label{tab_multivar_summary}",
             booktabs=T, linesep = "\\addlinespace",
             col.names = c(" ","Green spruce aphid abundance",
                        "Leaf-curling plum aphid first flight",
                        "Ceratium furca abundance","Methane flux")
             )%>%column_spec(2:5,width="2.5cm")%>%
             column_spec(1, width = "7cm")%>%kable_styling(latex_options = "basic",position = "center")
```

@@

Having demonstrated that non-N copula structure and asymmetric tail dependence are common
features of ecological data, we addressed Q2 by exploring via mathematical 
models two possible mechanisms of this structure which 
seem likely to operate in diverse ecological scenarios. The first mechanism 
is an extension of the well known Moran
effect, and was summarized conceptually in the Introduction. 
It explores the hypothesis that non-N copula structure and asymmetric tail dependence 
between environmental variables measured through time 
in two places may bring about similarly 
structured dependence between ecological variables in those places that are influenced by the 
environmental variables.

Let $E_i(t)$ be an ecological variable 
measured in location $i$ ($i=1,2$).
We initially assumed the variable follows AR(1) dynamics within a location, 
$E_i(t+1)=\beta E_{i}(t)+ \sqrt{1-\beta^2} \epsilon_i(t)$. 
We used $\beta=0.5$. Here the $\epsilon_i(t)$ represent environmental noise; they are 
standard-normal random variables that are independent for distinct times, $t$, but 
that were made to exhibit different kinds of 
dependence across locations in different simulations (see below).
The ecological variable, $E$, is very general, and could represent, for instance, deviations
of a population density from a carrying capacity or average density, deviations of
total plant community biomass or other community variable from an average value,
or flux of a biogeochemical variable such as methane or CO$_2$.

@@

For each value of $b$, we simulated the model once using $g_1$ and once using $g_2$, for 5000 time
steps in both cases, retaining $E_1(t)$ and $E_2(t)$ for the final 500 time 
steps of each simulation. We applied our model selection algorithms and 
non-parametric statistics, which were previously applied to 
data to answer Q1, to the 
simulation outputs to discover if the model could
give rise to non-N copula structure and asymmetric tail dependence in the 
ecological variable, in spite of the absence of such structure in the driver
environmental variables.

@@

For strong noise and using our nonlinear model (methods for Q2), 
copula statistics, generally speaking, were approximately similar 
between noise and model
outputs; however, similarity was reduced relative to the AR(1) 
and weak-noise cases, and, for a few simulations, tail dependence 
asymmetry statistics
even had opposite signs for noise and model outputs 
(Figs \ref{SM-fig_Cause4copula_C_kend_sRicker_nsd_1}-\ref{SM-fig_Cause4copula_SJ_kend_sRicker_nsd_1}). For instance, using a C copula with a large Kendall $\tau$,
$\cor_l - \cor_u$ was slightly positive for noise, but slightly negative for 
population model outputs (Fig. \ref{SM-fig_Cause4copula_C_kend_sRicker_nsd_1}); 
and likewise for SG and SJ copulas (Figs \ref{SM-fig_Cause4copula_SG_kend_sRicker_nsd_1}, 
ref{SM-fig_Cause4copula_SJ_kend_sRicker_nsd_1}). Dynamics were
strongly nonlinear for the model parameters we used, and nevertheless similarities
between noise and model-output copula structure, according to the statistics 
we used, were more noticable than discrepancies.
We again repeated the analysis with $r=1.3$. Discrepancies between noise and population
copula structure in this case were often glaring. Apparently noise of standard deviation 1
interacted very strongly with model nonlinearities when the model was in its overcompensatory 
regime.

Our model with environmental effects which differed in the
two tails could produce simulated ecological variables with visually apparent asymmetry 
of tail dependence and non-N copula structure, depending on the value of $b$ used. 
For $b=0.1$ (Fig. \ref{fig_ms_asym_b_0.1}) and $b=-0.1$ (Fig. \ref{SM-fig_ms_asym_b_-0.1}), 
for both $g_1$ and $g_2$,
asymmetric tail dependence and non-N copula structure were visually apparent.
Lower-tail (respectively, upper-tail) effects of the noise ($g_1$, resp., $g_2$) 
produced lower-tail (resp., upper-tail) dependence in the ecological variable.
Model selection results (Table \ref{SM-tab_summary_stat_asym_sens}) and
our nonparametric statistics (Tables \ref{SM-tab_stat_npa_asym}) 
statistically confirmed the visual result for $b=0.1, -0.1$. Results were more
complex for $b=0.9$ or $-0.9$, indicating that strongly autocorrelated (or strongly 
negatively autocorrelated) dynamics interact with the effect we have described.
Ecological-variable copulas were strongly non-N for $g_1$, but not for $g_2$ 
and asymmetries of tail dependence were not particularly noticable 
(Figs \ref{SM-fig_ms_asym_b_0.9}-\ref{SM-fig_ms_asym_b_-0.9} and Tables 
\ref{SM-tab_summary_stat_asym_sens}-\ref{SM-tab_stat_npa_asym}).
<!--***DAN: Shya, this has got to be an error. Results should be symmetric
with respect to the choice of g_1 and g_2 because the model is symmetric. Let's discuss.
I think there is likely a bug affecting Figs S51 and 51, and Tables S47-48, and 
possibly all results based on the same model.-->

@@

We addressed Q3 using both models and data. An initial 
idea was presented in the Introduction that
the distribution (through time) of a spatially averaged quantity should be influenced by the 
nature of the dependence between the local quantities being averaged, including their
copula structure and tail dependence. For instance, suppose $E_i(t)$ are ecological
variables measured at locations $i=1,\ldots,N$ and times $t=1,\ldots,T$, and suppose 
the distribution and temporal dynamics of the spatial average $\sum_i E_i(t)/N$ is of
interest. The $E_i(t)$ could be, for instance, local abundances of a pest or 
exploited species, or local fluxes of a greenhouse gas. If  $E_i$ and $E_j$ are 
strongly right-tail dependent for all or most location pairs $i$ and $j$, then
exceptionally large values tend to occur at the same time in all or most locations.
This should tend to increase the skewness of the distribution through time of the
spatial average. Similarly, left-tail dependence between local variables
will tend to decreases skewness of the spatial mean. Positive skewness of the 
spatial mean corresponds to the "spikiness" of the spatial-mean time series. 
The spatial-mean time series and its skewness may be quantities of 
principal importance for 
cases such as pest or resource abundance, for which extreme values (spikes) 
in the spatial-mean
quantity have disproportionately large effects.

@@

We also examined whether the nature of the
tail dependence of an environmental variable across space could influence the 
extinction risk of a metapopulation. We hypothesized that environmental noises
exhibiting greater left-tail dependence across different habitat patches would 
cause higher metapopulation extinction risks because then bad years for the 
component populations would occur simultaneously in many patches,
reducing rescue effects. Here, we are assuming that low values of the environmental
variable are "bad" for the populations being modelled, and high values are "good".
We provided a test of the conceptual validity of this hypothesis using a metapopulation
extension of the Lewontin-Cohen model,
$\vec{P}(t+1) = D \lambda(t) \vec{P}(t)$,
where $\vec{P}(t)$ is a vector of population densities, 
the $i^{th}$ component $P_i(t)$ representing
the density in the $i^{th}$ habitat patch ($i=1,\ldots,N$) at time $t$. 
The $N \times N$ matrix $\lambda(t)$ was diagonal with $i^{th}$ diagonal entry 
$\exp(r+\epsilon_i(t))$. Here $r$ is a growth rate; we used $r=0$. The
$\epsilon_i(t)$ represent environmental noises in each patch. They were standard-normally 
distributed in each patch, were independent through time, and showed the same 
spatial correlations (to within sampling error) for every simulation, 
but were made to exhibit right- or left-tail dependence between patches in
different simulations (Appendix \ref{SM-ext_risk_model}). The $N \times N$
matrix $D$ was a dispersal matrix. For simplicity, we considered habitat patches 
to be arranged in a line, with dispersal occurring either to nearest neighbors only,
or globally, in different simulations. The dispersal 
rate was $d$ (Appendix \ref{SM-ext_risk_model}). After each time step, 
if the population density in a patch was below $1$ (an extinction threshold), it was
set to $0$. We simulated the model 10000 times for each combination of parameters,
starting from initial populations $p_0 = 50$ in each patch, 
and calculated extinction risk after $25$ time steps.
